Examining readability of text generated through GPT-3

### 2023-04-02 paper/paper.tex
update paper

\begin{document}
\maketitle
\begin{abstract}
    This paper examines readability-related metrics of AI-generated text, specifically texts produced by OpenAI's GPT-3. Generation was performed for the English language across 10 topics and the 6 proficiency levels defined by the CEFR framework, and outputs were compared with Simple Wikipedia and (standard) Wikipedia articles on the same topics. 7 commonly used complexity metrics have been employed, targeting chiefly lexis and syntax. In the limited scope of this experiment, GPT-3 is observed to be at least partially aware of what constitutes reading difficulty.
\end{abstract}

@@

This work aims to make a contribution to these questions by verifying the readability level of text content generated by GPT-3 with regard to 10 topics and the 6 proficiency levels defined by the CEFR framework. A number of complexity metrics are computed on the generated texts and compared with Simple Wikipedia\footnote{\url{https://simple.wikipedia.org/wiki/Main_Page}} as well as Wikipedia\footnote{\url{https://www.wikipedia.org/}} articles on the same topics in order to gauge the true readability levels of the model outputs.
\section{Dataset and metrics}
\subsection{Dataset}

@@

@@

@@

Since recent research (e.g. \citealp{Deutsch_2020}) has also focused on readability assessment through neural approaches without such metrics, this could also be explored with AI text data. An attempt was made in relationship to this experiment to train a readability analyzer with GLoVe word embeddings \citep{pennington2014glove} and Wikipedia data, but this unfortunaly was inconclusive, and thus results from this analyzer haven't been reported.

Mainly, however, future extensions of this experiment or related experiments would benefit from higher volume. Simply put, more data is necessary across multiple domains to make educated conclusions about the quality of GPT-3 outputs in a teaching setting. While data is easy to generate, quality annotations on it are yet unavailable. Indeed, it is a good question whether making any is a fruitful enterprise, as the actual models behind GPT-3 and ChatGPT are constantly updated - thus making it hard to pinpoint what exactly is being studied.

@@

>>>>

\begin{document}
\maketitle
\begin{abstract}
    This paper examines readability-related metrics of AI-generated text, specifically texts produced by OpenAI's GPT-3. Generation was performed for the English language across 10 topics and the 6 proficiency levels defined by the CEFR framework, and outputs were compared with Simple Wikipedia and (standard) Wikipedia articles on the same topics. 7 commonly used complexity metrics have been employed, targeting chiefly lexis and syntax. In addition, a classifier is also trained on Wikipedia data and used to evaluate the generated texts. In the limited scope of this experiment, GPT-3 is observed to be at least partially aware of what constitutes reading difficulty.
\end{abstract}

@@

This work aims to make a contribution to these questions by verifying the readability level of text content generated by GPT-3 with regard to 10 topics and the 6 proficiency levels defined by the CEFR framework. A number of complexity metrics are computed on the generated texts and compared with Simple Wikipedia\footnote{\url{https://simple.wikipedia.org/wiki/Main_Page}} as well as Wikipedia\footnote{\url{https://www.wikipedia.org/}} articles on the same topics in order to gauge the true readability levels of the model outputs.
\section{Dataset and methods}
\subsection{Dataset}

@@

\subsection{Text classifier}

A classifier was trained on Wikipedia and Simple Wikipedia data, specifically on a train dataset of approximately 240k paragraphs, evenly distributed across sourced from Wikipedia and Simple Wikipedia articles. A test set of 40k entries, sourced and distributed in the same manner as the train set, was also obtained to evaluate the data. The task of the classifier was to determine whether a given paragraph had originally been a part of Simple Wikipedia or (standard) Wikipedia - $P(standard)$ is the derived readability metric reported in the results.

The model takes BERT sentence embeddings as input, which are fed through a single-layer LSTM with hidden size 126. LSTM outputs are then used to compute a distribution over the two labels in the dataset, namely "simple" (paragraph obtained from Simple Wikipedia) and "standard" (paragraph obtained from standard Wikipedia). Training was performed with the Adam optimizer and a batch size of 50 for 5 epochs. After evaluating the model at each epoch, it was observed that after the third epoch there are signs of overfitting, therefore the final parameters were those obtained after 3 epochs of training. Hardware-wise, the free tier of Google Colab was used - do to time limits set on the platform, several days were necessary for the full training process. 
After the third epoch, the model achieves 87.8\% macro-averaged $F_1$-score on the test set. While this is not state-of-the-art performance, it is likely good enough for the scope of this project.

@@

\subsection{Classifier scores}

\input{tables/classifier_scores.tex}

Table~\ref{tab:classifier} shows classifier outputs, with highlighted values for each topic and difficulty level associated to lowest readability. The Wikipedia entries that were reported on in previous sections are not included here, as they may have been part of the train set of the classifier.

These results show again that GPT-3 has learned what constitues readability, at least to some extent. For 4 of the topics presented in this study, the C2 prompt achieves the highest complexity (or lowest readability) level, with 3 more having B2 as the top entry.

As was observed with the metrics, there tends to be good differentiation between the extremes (i.e. between A1 and C2 texts), but this is less true in the middle portions. Interestingly, prompts for C1 level texts often do not yield particularly difficult texts according to the classifier. Such a claim is confirmed by looking at the average values for each level, where C1 does score higher than A1 and A2, but lower than B2. Whether this is caused by this classifier's failure to capture some aspect of text readability, ot by GPT-3's lessened ability to generate material for this target level, remains to be confirmed by more robust experiments.

On the matter of inherent topic complexities, there is unfortunately not much to talk about - no topic or set of topics emerge as consistently more difficult than others. However, \emph{Economics} and \emph{Semantics} do get the highest average score, indicating they are the least readable, while (more surprisingly) \emph{Color Blindness} is the most readable on average.

@@

Since recent research (e.g. \citealp{Deutsch_2020}) has also focused on readability assessment through neural approaches without such metrics, this paper also attempted to adopt this strategy by training a classifier on Wikipedia and Simple Wikipedia data. Outputs derived from this model also support the theory of GPT-3 being at least somewhat aware of readability-related aspects of language. With regards to this section, it is nonetheless paramount to utilise a larger, SOTA readability model for exploring complexity - as the one employed here is only suitable for prototyping - as well as to perform the experiment on a richer repository of GPT-3 data.

In summary, future extensions of this experiment or related experiments would most benefit from higher volume. Simply put, more data is necessary across multiple domains to make educated conclusions about the quality of GPT-3 outputs in a teaching setting. While data is easy to generate, quality annotations on it are yet unavailable. Indeed, it is a good question whether making any is a fruitful enterprise, as the actual models behind GPT-3 and ChatGPT are constantly updated - thus making it hard to pinpoint what exactly is being studied.

@@

I thank Prof. Dr. Walt Detmar Meurers for exposing me to the topics of readability and automatic readability assessment, and for the excellent course he taught on the matter.
